---
title: "README.Rmd"
author: "Arwenn Kummer"
date: "2025-02-24"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GIS in R

In this exercise I will be mapping the current distribution of three species of protea chafer, to determine if their current ranges overlap with endangered ecosystem types (using iNat and SANBI/CapeNature data). I aim to create an interactive map using mapview() that allows viewers to pan, zoom and inspect individual observations, the ecosystem threat status that they occur in, as well as the name of the ecosystem the occur in.

## Basemaps and Data

*Current ecosystem/vegetation types sourced from City of Cape Town (CoCT)*: while working with the data, it became apparent that this data set is redundant, as the relevant information I would have needed from it is already encased within the SANBI/CapeNature data set that I used for the Threatened Ecosystem areas.

*Threatened ecosystem types sourced from SANBI/CapeNature*: <https://bgis.sanbi.org/SpatialDataset/Detail/611>. This data maps the current threat statuses for various ecosystms in the Western Cape.

*Species observation data sourced from iNaturalist*: <https://www.inaturalist.org/observations?place_id=6986&subview=map&taxon_id=423700&verifiable=any&view=species&iconic_taxa=Insecta>. This data includes research grade, wild observations, though some observations occur within Kirstenbosch Botanical Gardens, and thus may be considered 'captive'. Sub-species are included in the broader species classification for ease.

### Loading Data files into R:

#### Threatened ecosystems in the Western Cape

This data was sourced from the 2016 Ecosystem Threat Status project conducted by SANBI and CapeNature (<https://bgis.sanbi.org/SpatialDataset/Detail/611>).

```{r}
# load the sf library - this allows us to work with shape files.
library(sf)

# read the shapefile into R
threat_eco <- st_read("C:/Users/Arwenn Kummer/Documents/GIT/GISinR/WCBSP_Ecosystem_Threat_Status_2016/BSP_Ecosystem_Threat_Status_2016.shp")
view(threat_eco) # view the data to determine which variables are present and what information is provided. 

# determine the coordinate reference system
st_crs(threat_eco)

```

```{r}
# load the tidyverse library - gives us access to numerous functions, including ggplot()
library(tidyverse)

# plot threat_eco in ggplot() to do a visual check. The `CNa1_ETS14` variable gives us the threat status of each ecosystem
ggplot() +
  geom_sf(data = threat_eco, aes(fill = `CNa1_ETS14`)) +
  theme_minimal()
```

## Observational Data from iNaturalist

Three of the most common protea chafers found in the Western Cape and South Africa will be used in this distribution mapping exercise; Cape Protea Chafer (Tricostetha capensis), Green Protea Chafer (T. fascicularis) and Signal Protea Chafer (T. signata). These will be read into R using a function from rinat(): get_inat_obs(). Each species' observations will be stored in a data frame and subsequently converted to a shapefile for mapping.

#### Cape Protea Chafer:

```{r}
# load the rinat() library - allows us to read data from iNaturalist directly
library(rinat)

# load observations
cpc <- get_inat_obs(taxon_name = "Trichostetha capensis",
                   bounds = c(-35, 17, -30, 25), # confined searches to the Western Cape region
                   maxresults = 400)
head(cpc)
```

```{r}
# select only research grade and wild observations
cpc <- cpc %>% filter(positional_accuracy<46 & 
                      latitude<0 &
                      !is.na(latitude) &
                      captive_cultivated == "false" &
                      quality_grade == "research")

class(cpc)

# convert the data frame to a spatial object to allow mapping
cpc_sf <- st_as_sf(cpc, coords = c("longitude", "latitude"), crs = 4326)

# check the coordinate reference system
st_crs(cpc_sf)
```

```{r}
plot(cpc_sf) # plot the shape file to check that everything works and runs
```

#### Green Protea Chafer:

```{r}
# load observations
gpc <- get_inat_obs(taxon_name = "Trichostetha fascicularis",
                    bounds = c(-35, 17, -30, 25), # confined searches to the Western Cape region
                    maxresults = 300)

head(gpc)
```

```{r}
# select only research grade and wild observations
gpc <- gpc %>% filter(positional_accuracy<46 & 
                      latitude<0 &
                      !is.na(latitude) &
                      captive_cultivated == "false" &
                      quality_grade == "research")
class(gpc)

# convert the data frame to a spatial object
gpc_sf <- st_as_sf(gpc, coords = c("longitude", "latitude"), crs = 4326)

# check the coordinate reference system
st_crs(gpc_sf)
```

```{r}
plot(gpc_sf) # plot the shape file to check that everything works and runs

```

#### Signal Protea Chafer:

```{r}
# load observations
spc <- get_inat_obs(taxon_name = "Trichostetha signata",
                    bounds = c(-35, 17, -30, 25),
                    maxresults = 120)
head(spc)
```

```{r}
# select only research grade and wild observations
spc <- spc %>% filter(positional_accuracy<46 & 
                      latitude<0 &
                      !is.na(latitude) &
                      captive_cultivated == "false" &
                      quality_grade == "research")

class(spc)

# convert the data frame to a spatial object
spc_sf <- st_as_sf(spc, coords = c("longitude", "latitude"), crs = 4326)

# check the coordinate reference system
st_crs(spc_sf)
```

```{r}
plot(spc_sf) # plot the shape file to check that everything works and runs
```

We can see that the data for threatened ecosystems and each of the species plots correctly. Now we can start layering them and working on an interactive map

# Mapping

## Beetle distribution and Threatened Ecosystems in the Western Cape

First we'll create a static map that shows the ecosystem threat-status, with the beetles distribution shape files overlaid. We'll do this using ggplot() for now.

```{r}
# Let's make the map a bit prettier: plotting the ecosystem threat status with a different colour scheme
ggplot() +
  geom_sf(data = threat_eco, aes(fill = `CNa1_ETS14`)) +
  scale_fill_manual(values = c("#ffac27", "#e7815d", "#ce5693", "#b62bc9", "#9d00ff")) +
  labs(fill = "Ecosystem threat status")
  theme_minimal()
```

I scoured Google and Stackexchange trying to find out how to add a legend for the beetles but couldn't find a solution that ran. As such the legend is here in plain text: T. capensis = pink T. fascicularis = light green T. signata = sky blue

```{r}
# Overlap the beetles' distributions
  geom_sf(data = threat_eco, aes(fill =  `CNa1_ETS14`)) +
  scale_fill_manual(values = c("#ffac27", "#e7815d", "#ce5693", "#b62bc9", "#9d00ff")) +
  labs(fill = "Ecosystem threat status")+
  geom_sf(data = cpc_sf, color = "pink", size =0.8) +
  geom_sf(data = gpc_sf, color = "lightgreen", size = 0.8) +
  geom_sf(data = spc_sf, color = "skyblue", size = 0.8) +
  theme_minimal()


```

This is all well and good, and provides us with a broad scale view of the species distributions across the various ecosystems. But we want to be able to look closer and see exactly where these species fall, and in what ecosystem type. To achieve this, we will generate an interactive map.

# Interactive Map!

```{r}
library(mapview) # allows us to create an interactive map
library(RColorBrewer) # loads in a palette of colour schemes

mapview(threat_eco, zcol = c("CNa1_ETS14"), col.regions = brewer.pal(5,"RdYlBu"), map.types = "CartoDB.Positron", layer.name = c("Ecosystem Threat Level")) + 
  mapview(threat_eco, zcol = c("NAME"), layer.name = c("Ecosystem name"), hide = TRUE, legend = FALSE) +
  mapview(cpc_sf, col.regions = "hotpink", layer.name = c("Trichostetha capensis")) + 
  mapview(gpc_sf, col.regions = "green", layer.name = c("Trichostetha fascicularis")) + 
  mapview(spc_sf, col.regions = "turquoise", layer.name = c("Trichostetha signata"))
```

Note that the `ecosystem names` layer is hidden when the map generates, but this can be toggled on and off using the layer icon below the zoom buttons in the top left hand corner. Mousing over the map will show the ecosystem names. The legend has been hidden for visibility purposes.

Now we are able to visually assess the current distribution of these species, and which ecosystem types they fall in. Further more, we can see that most fall on the boundary of a threatened ecosystem, or within a threatened ecosystem. This information can be used to inform conservation efforts and to determine which species and subspecies are most at risk from occurring within these threatened ecosystems.
